---
title: R Tutorial
order: 3
---

## Time Series Autocorrelation in R
### Informational Videos on Autocorrelation

1. Watch Introduction to exploring autocorrelation
<iframe width="560" height="315" src="https://www.youtube.com/embed/iSfRh8nxAgI" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

2. Watch Autoregressive vs. moving average processes
<iframe width="560" height="315" src="https://www.youtube.com/embed/KXsYYhdkjgs" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

---

Last week we talked about the different time scales as one type of signal that 
was embedded in a timeseries. This week we're going to talk about a different 
issue with time series. 

Let's start by thinking about an extreme case of a time series. 

[INSTRUCTOR:DRAW ON BOARD NORMAL DISTRIBUTION CENTERED AT ZERO]

Say I have a normal distribution centered on zero and at each time step I 
randomly draw a value from the distribution. If I plotted this time series, 
what do you think it would look like? 

[INSTRUCTOR: HAVE THEM WALK YOU THROUGH DRAWING THE TIME SERIES. The point to
make is that the choice of the value at time t is not dependent on its value at t-1]

When a time series is generated this way -- where the past has no influence on 
the value in the future -- we call is white noise. We'll make a white noise 
time series by doing just what we said. We'll pull 273 values randomly from a 
normal distribution, which R already has a function for generating. We'll just 
use the default values of a mean of zero and sd=1

```{r}
library(fpp3)
set.seed(123)
whitenoise = tsibble(sample=1:273, wn = rnorm(273, mean=0.18), index=sample)           
autoplot(whitenoise, wn) + geom_hline(yintercept=.18)
```
How biological does this look to you?

Now, let's look at the NDVI timeseries from portal that we were working with last 
week. So, just like last week, we'll load the data, load the data into 
a tsibble. And let's plot it to compare to the white noise.

```{r}
data = read.csv('portal_timeseries.csv')
data = mutate(data, month = yearmonth(date))
data_ts = as_tsibble(data, index=month)
autoplot(data_ts, NDVI) + geom_hline(yintercept=.18)

```

Visually, how do these look different? Use that we learned last week on 
time series decomposition to help inform why these look different.

[Instructor: We want them to think about the different time scales in the NDVI 
data, which creates some memory or correlations between values and makes the
values not truly independent of each other]

The value at one time step is not necessarily independent from previous time 
steps. We can explore that dependence by looking at lag plots. How much does 
one time step tell you about the next?

Lag plots are simply the correlation between values at time t and some time 
step in the past, where the lag is the amount of time between the two time points. 

```{r}
gg_lag(data_ts, NDVI, geom= "point") + labs(x= "lagged NDVI")
```
The lag.plot is showing you correlation within a time series, called 
autocorrelation. So in our top left panel, we have each data point plotted 
against the value one month prior, where the current or time t data point is 
on the y and the value one month ago is on the x. These plots go up to a 9 month 
lag. 

What do you see from these plots?

[Instructor: In lag one we see some correlation between this month and the 
previous month, and that signal breaks down over time. The different colors 
are the different months, which helps visualize if there are specific months 
that are outliers from the overall relationship. The light green dots often
end up in the upper left corner -- high values at time t but lot values at
t-1. This is because in the summer plant productivity erupts as soon as the
summer monsoon season arrives, so the landscape is usually pretty brown in june
but green in july]

We can quantify these relationships by calculating a correlation coefficient 
at each one of these lags. An autocorrelation function or ACF or correlogram 
conducts those correlations.

```{r}
ACF_results = ACF(data_ts, NDVI, lag_max=12)
ACF_results
```

As our lag plots suggested, our strongest correlation is at 1 month and the 
signal decays over time. ACF results are most often presented as graphs, so lets
look at a typical ACF plot:

```{r}
autoplot(ACF_results)
```
Our x-axis is the lag value in months. We see the strongest correlation is at a 
lag of 1 month, and then the autocorrelation in the time series drop off 
substantially. We get little blips every 12 months which is the seasonality 
coming through and perhaps a weak signal of a negative relationship around 4-8 month lags. 

The blue lines plotted are the 95% confidence interval for the autocorrelation 
of a time series (+- 1.96/(sqrt(T))), where T is the length of the time series. 
If more than 5% of spikes are outside this bound, your time series is probably 
not white noise. 

Let's see what white noise looks like in one of these graphs

```{r}
gg_lag(whitenoise, wn, geom= "point") + labs(x= "lagged value")

```
No relationship between the points, which is what we would expect sicne we deliberately pulled them without any dependence of one time point on the next. Let's look at the acf plot for this data

```{r}
ACF(whitenoise, wn, lag_max=12) |> autoplot()
```

No correlations go above the blue lines. You can sometimes get a spike edging 
above the blue lines even with white noise. Sometimes randomess can result in 
correlations but they should not be very strong. 

If there's a pattern to your spikes, that is usually another good sign that you
have autocorrelation structure in your time series

[INSTRUCTOR: Have students make lag and ACF plots for precip and rodents. 
Tell them to do PPT first then rodents, their reaction to the rodent plot
will be a signal for when they've finished the exercise]

```{r}
gg_lag(data_ts, rain, geom= "point") + labs(x= "lagged value")
```
```{r}
ACF(data_ts, rain, lag_max=12) |> autoplot()
```
```{r}
ACF(data_ts, rodents, lag_max=12) |> autoplot()
```
```{r}
gg_lag(data_ts, rodents, geom= "point") + labs(x= "lagged value")
```
The slow decay in the ACF graph can indicate you have a trend in your data 
because if your time series is increasing over time, this month will be higher 
and next month will be higher and the lonth after that will be higher, which 
will generate correlations between your time points, not because the months 
are correlated but because of the trend. In time series analysis language your
time series is non-stationary and many time series analyses would want you to 
detrend you data. So, let's look at the decomposition for our rodent data

```{r}
stl_rodents = data_ts |> model(STL(rodents ~ trend(window=21) + season(window = 13), robust = TRUE)) |> components() 
autoplot(stl_rodents)
```
We have long-term wiggles, but not really a strong trend. Another possibility 
is the echo of time periods past.

MAKE TIME SERIES GRAPH ON BOARD

If Y_t and Y_t-1 are strongly correlated and Y_t-1 and Y_t-2 are strongly 
correlated then presumably Y_t and Y_t-2 are correlated even if there is no 
causal factor involved. We can examine this by using a partial acf. The partial 
acf examines the correlations at a specific lag value, controlling for the lags 
at all shorter lags. 

We're going to use one of the functions in fpp3 that will give us both the ACF 
and the PACF graphs for the rodents.

```{r}
gg_tsdisplay(data_ts, rodents, plot_type=c("partial"))
```
HAVE STUDENT DO THIS FOR NDVI AND RAIN

```{r}
gg_tsdisplay(data_ts, NDVI, plot_type="partial")
```
```{r}
gg_tsdisplay(data_ts, rain, plot_type="partial")
```

The rodents show a classic signal of an autoregressive model. 
An autoregressive model is when the value at time t depends on the values at 
previous time steps. 

A classic autoregressive model is a random walk. A random walk is where the 
value at time t is dependent on the value at time t-1. We'll make a random walk  
series by grabbing a random value from the normal distribution and then adding 
the value from the previous time step. This creates the AR1 signal where 
there is a strong 1 time step autocorrelation in the time series.


```{r}
set.seed(123)
x = w = rnorm(1000)
for (t in 2:1000) x[t] = x[t-1]+w[t]
randomwalk = tsibble(time=1:1000, rwalk = x, index=time)
gg_tsdisplay(randomwalk, plot_type="partial")

```

you can use these plots to get a better understanding of your time series.
What dynamics are occurring? What questions might be worth asking or what is 
surprising to you that doesn't seem to be coming through? If you are doing time
series modelling you can also use these graphs to help inform what structure
of a time series models makes the most sense given your data. I think Ethan
will be talking more about that next week.

Understanding your autocorrelation structure is also important statistically
Regression approaches with autocorrelated data will give underestimates of the 
variance inflated test statistics, and narrow CIs. 

[Instructor: If time, you can show students crss-correlations because they
generally think it's useful and interesting, but it is not necessary for the
course]

Finally, sometimes you want to understand how two time series are correlated
across different lags. You can use the cross-correlation function to dig into that

```{r}
ccf.plantsrain = ccf(data_ts$rain, data_ts$NDVI)
```
These do both forwards and backwards lags. The first series listed is the one 
that gets lagged. In this case, we have reason to believe that PPT from the past 
influences NDVI in the future. 

Some take home messages for autocorrelation: 


1. Autocorrelation is useful in that information about the past and future states 
of the system are encoded in the timeseries. This is information that can 
potentially be leveraged for forecasting.
2. Autocorrelation is a statistical pain in the butt. Statistical approaches
Assume iid: independent and identically distributed errors. i.e. that your
data is a random draw from an underlying distribution. But autocorrelation
means that your data is not a random draw. Each draw is influenced by the
previous draw. This means that if you put a time series through a regression 
your confidence intervals will be smaller than they should be. Parameter 
estimates are generally ok. Need to deal with that autocorrelation for 
statistical tests many modern statistical approaches in R have a method for 
dealing or specifying autocorrelated errors - sometimes referred to as 
covariance in the errors.

### Old R tutorial Videos on Autocorrelation

1. Watch Generating random time series in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/iO3yXnnSksM" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

2. Import portal_timeseries.csv in R using read.csv. 
Convert the NDVI data column into a time series object using ts(). Name that time series object: NDVI.ts
Convert the rodent column into a time series object using ts(). Name that time series object: rats.ts 
Convert the rain column into a time series object using ts(). Name that time series object: rain.ts
(Why am I telling you what to name these things? So that when I type code in the video, it is easier for you to follow along. That's all!)

3. Watch Exploring autocorrelation through lag plots
<iframe width="560" height="315" src="https://www.youtube.com/embed/V-4W8Wl-H9E" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

4. Generate a lag plot for rats.ts using lag.plot() 

5. Watch Using acf() in R to explore autocorrelation
<iframe width="560" height="315" src="https://www.youtube.com/embed/deV2SlJ2Dnw" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

6. Generate an acf plot for rain.ts using acf()
Generate an acf plot for rats.ts using acf()

7. Watch Using pacf() to diagnose time lags for autoregressive models
<iframe width="560" height="315" src="https://www.youtube.com/embed/4W9uI4pjc8U" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

8. Using tsdisplay() examine the acf and pacf for NDVI.ts
Using tsdisplay() examine the acf and pacf for rain.ts

9. Watch Exploring lagged correlations between different time series
<iframe width="560" height="315" src="https://www.youtube.com/embed/dKDgihvtZGk" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
