---
title: "Species Distribution Model Demo"
output: html_document
---

## Species Distribution Models

* Where species are expect to occur
* Can be based on lots of things, but often just environmental variables
* Very simple version - do not use this in production

## Load some packages

* `dismo` is the major package for SDM in R

```{r}
library(ggplot2)
library(dismo)
library(dplyr)
```

## Data

* Two kinds of data for SDMS
* Locations for where a species occurs
    * And ideally where it is absent
    * When no absences make fake absenses called "background"
* Spatial data on predictor variables
    * Most commonly temp & precip based
    * But that is mostly convenience based, not biology based
    * Need both current and future (forecast) values
* Location data
    * Breeding Bird Survey of North America on Hooded Warbler
* Environmental data
    * Temperature (minimum) & precipitation
    * CMIP5 50 year forecasts (coupled model intercomparison project)
* Download data from [https://course.naturecast.org/data/sdm_data.zip](https://course.naturecast.org/data/sdm_data.zip)

```{r}
hooded_warb_data = read.csv("hooded_warb_locations.csv")
env_data_stacked_current = stack("env_current.grd")
env_data_stacked_forecast = stack("env_forecast.grd")
head(hooded_warb_data)
plot(env_data_stacked_current)
```

## Determine environment where species is (and isn't)

* Combine the two data types
* Info on the environment where a species occur and where it doesn't

```{r}
hooded_warb_locations = select(hooded_warb_data, lon, lat)
hooded_warb_env = data.frame(extract(env_data_stacked_current, hooded_warb_locations))
hooded_warb_data = cbind(hooded_warb_data, hooded_warb_env)
```

* Can look at this in a 2D case

```{r}
ggplot(hooded_warb_data, aes(x = tmin, y = precip, color = present)) +
  geom_point()
```
* Find regions of climate space where the species is likely to occur
* This species is associated with high temps and precips

## Modeling species distribution

* Many different ways to model probability of species presence
* Look at one of the simplest
    * Generalized linear modeling
    * Multivariate logistic regression
    
```{r}
logistic_regr_model <- glm(present ~ tmin + precip,
                           family = binomial(link = "logit"),
                           data=hooded_warb_data)
summary(logistic_regr_model)
```

## Evaluate the model performance

```{r}
presence_data = filter(hooded_warb_data, present == 1)
absence_data = filter(hooded_warb_data, present == 0)
evaluation <- evaluate(presence_data, absence_data, logistic_regr_model)
plot(evaluation, 'ROC')
```

* Looks good, but is biased when using absenses from large scales

## Plot the model predictions

* Use the overloaded `predict` function from the `raster` package
* Arguments
    * Raster of environmental conditions
    * Model
    * `type = "response"` to get probabilities

```{r}
predictions <- predict(env_data_stacked_current, logistic_regr_model, type = "response")
present <- select(filter(hooded_warb_data, present == 1), lon, lat)
plot(predictions, ext = extent(-140, -50, 25, 60))
points(present, pch='+', cex = 0.5)
plot(predictions > 0.5, ext = extent(-140, -50, 25, 60))
points(present, pch='+', cex = 0.5)
```

* Can also choose a different threshold to give better classification
* E.g., > 25% probability of occuring
* Can choose this automatically

```{r}
tr <- threshold(evaluation, 'prevalence')
plot(predictions > tr, ext = extent(-140, -50, 25, 60))
points(present, pch='+', cex = 0.5)
```
## Make forecasts

* Use forecast data for future environmental conditions
* CMIP5

```{r}
predictions <- predict(env_data_stacked_forecast, logistic_regr_model, type = "response")
plot(predictions, ext = extent(-140, -50, 25, 60))
plot(predictions > 0.5, ext = extent(-140, -50, 25, 60))
plot(predictions > tr, ext = extent(-140, -50, 25, 60))
```

