---
title: "Assignment Solution: Time series modeling in R"
subtitle: Instructor answer key -- not published to the course website
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  enabled: true
  warning: false
  message: false
---

::: {.callout-warning}

This file is excluded from the site render in `_quarto.yml` so that it is never
published alongside the assignment. Render it by hand when you want to refresh it:

```sh
quarto render lessons/R-time-series-modeling-3/assignment_solution.qmd
```

Unlike the rest of the site, this document **executes** its R code, so it needs
`tsibble`, `fable`, `feasts`, `ggtime`, and `dplyr` installed.

:::

```{r}
#| label: setup
library(tsibble)
library(fable)
library(feasts)
library(ggtime)
library(dplyr)
```

## Part 1: Look at the data

```{r}
#| label: load
barco <- read.csv("../../data/barco_oxygen_timeseries.csv") |>
  mutate(month = yearmonth(date)) |>
  as_tsibble(index = month)
```

```{r}
#| label: tsdisplay
gg_tsdisplay(barco, oxygen)
```

## Part 2: White noise

```{r}
#| label: mean-model
avg_model <- model(barco, MEAN(oxygen))
report(avg_model)
```

```{r}
#| label: mean-fit
avg_aug <- augment(avg_model)
autoplot(avg_aug, oxygen) +
  autolayer(avg_aug, .fitted, color = "red")
```

```{r}
#| label: mean-resid
gg_tsresiduals(avg_model)
```

## Part 3: Autoregressive models

```{r}
#| label: ar-model
ar_model <- model(barco, AR(oxygen ~ order(2)))
report(ar_model)
```

```{r}
#| label: ar-fit
ar_aug <- augment(ar_model)
autoplot(ar_aug, oxygen) +
  autolayer(ar_aug, .fitted, color = "orange")
```

```{r}
#| label: ar-resid
gg_tsresiduals(ar_model)
```

## Part 4: Seasonal ARIMA

```{r}
#| label: arima-model
arima_model <- model(barco, ARIMA(oxygen))
report(arima_model)
```

```{r}
#| label: arima-fit
arima_aug <- augment(arima_model)
autoplot(arima_aug, oxygen) +
  autolayer(arima_aug, .fitted, color = "orange")
```

```{r}
#| label: arima-resid
gg_tsresiduals(arima_model)
```

## Part 5: Adding covariates with TSLM

```{r}
#| label: tslm-temp
tslm_temp <- model(barco, TSLM(oxygen ~ temperature))
report(tslm_temp)
```

```{r}
#| label: tslm-both
tslm_model <- model(barco, TSLM(oxygen ~ temperature + chla))
report(tslm_model)
```

```{r}
#| label: tslm-fit
tslm_aug <- augment(tslm_model)
autoplot(tslm_aug, oxygen) +
  autolayer(tslm_aug, .fitted, color = "orange")
```

```{r}
#| label: tslm-resid
gg_tsresiduals(tslm_model)
```

## Part 6: Dynamic regression

```{r}
#| label: arimax-temp
arimax_temp <- model(barco, ARIMA(oxygen ~ temperature))
report(arimax_temp)
```

```{r}
#| label: arimax-both
arimax_model <- model(barco, ARIMA(oxygen ~ temperature + chla))
report(arimax_model)
```

```{r}
#| label: arimax-fit
arimax_aug <- augment(arimax_model)
autoplot(arimax_aug, oxygen) +
  autolayer(arimax_aug, .fitted, color = "orange")
```

```{r}
#| label: arimax-resid
gg_tsresiduals(arimax_model)
```

## Part 7: Wrap-up
