---
title: R Tutorial
order: 3
---

## Time Series Decomposition in R

### Written version of the lesson. It will vary from the videos (below), but the 
### core info is the same. You may find it helpful to watch the videos without 
### "in R" in their title as these are more informative videos with visuals that are not  ### present in the written document.

R packages for the lesson:
fpp3
ggplot2
slider

## Inherent scales within a time series

Pull up the Mauna Loa CO2 time series (https://www.climate.gov/news-features/understanding-climate/climate-change-atmospheric-carbon-dioxide). 

What do you see? 

[Instructor: point out the two patterns in the data - the long-term trajectory 
and the squiggle which is the seasonal signal. Zoom in and ask if all the 
seasonal oscillations are the same. Point out that what we're seeing is change
on multiple time scales - Long-term (across years), seasonal (within a year),
and random variation or impacts from stochastic events.

To pull these scales out of the data so we can see them better or analyze them, 
we can use an approach called time series decomposition. The output from time 
series decomposition can be used for analyses - for example, seasonally-adjusted 
data is data that has gone through some form of time series decomposition to pull 
out the seasonal signal. Even if you don't need seasonally adjusted data, time 
series decomposition can be a good data visualization approach for better
understanding your time series and whats going on and what you might need to 
think about.

There are a variety of time series decomposition approaches, but they all boil 
down to 3 basic steps.
1) we fit something to the observed data to extract the trend.
2) we fit a seasonal model to the remaining data to pull out the season
3) whatever is left over is the irregular fluctuations (residuals)

## Loading the data
Let's read in our data:

```{r}
data = read.csv('portal_timeseries.csv', stringsAsFactors = FALSE)
head(data)
```

This is a data file that we will be using for the next few weeks. This data is 
from my field site in Arizona. At my field site, called the Portal Project, we 
monitor rodents, plants, and weather. I've given you data here for each of these 
components: rodent abundance, precipitation amount in mm, and the plant data 
I've given you is actually NDVI, which is a measure of greenness that comes from 
satellites. This data is given in a format where each data point is associated 
with a specific month. Our data is not collected exactly a month apart, but this 
approach requires regularly collected data, so I've fudged things to make this 
data fit the constraints of this approach. You'll see that I have a date column 
that is not in the international date format. The first column is just a row 
index that I forgot to suppress when I made this file.

Let's format the date column but do not turn it into a tsibble yet.

[Instructor: wait while they do this without you]
```{r}
data = mutate(data, date=lubridate::mdy(date))
```

## fpp3

We're going to be working with the fpp3 package which was written by forecasters 
to help teach and conduct forecasting. fpp3 is a metapackage, which is a package 
that bundles multiple packages together. Ethan may have you work with the 
individual packages more directly, but we're going to start by letting fpp3 make 
things easy for us as we learn. 

So let's load the package

```{r}
library(fpp3)
```

fpp3 needs our data in a tsibble format. I taught you last week how to put data 
in a tsibble using the date as the index but we're going to convert our dates into 
just a year and a month. So instead of March 15th 1992, it'll just be March 1992. 
You may wonder why year month and not into a date? Many of time series analyses 
require regularly collected data - monthly, daily, quarterly, annually. Our data 
is collected roughly monthly, so we're dropping the day information and turning 
it into a monthly time series.
```{r}
data = mutate(data, month=tsibble::yearmonth(date))
class(data$month)
```
So we're dealing with a different type of date object - the yearmonth, which is 
not a date object but works in a similar way but with months and years. Let's 
see how this looks in our dataframe.

```{r}
head(data)
```

Now we're ready to turn our data into a tsibble where the index isn't date and month

```{r}
data_ts = as_tsibble(data, index=month)
head(data_ts)
```

Let's look at some of our data by plotting it. 

```{r}
library(ggplot2)
ggplot(data_ts, aes(month, NDVI)) +
  geom_point() +
  geom_line()

```
As you can see, the greenness at this desert site varies over time and it does so
in a couple of ways. Each dot is a month. If we focus on going from month to month
we see periods where there are less change and periods where there are big jumps. 
If we step back and focus on the lines which connect our monthly data, we see 
a pattern of peaks and troughs. And if we step back even further, we see that
years differ in the magnitude of those peaks.

In every time series, there will be multiple levels or scales of information 
- the variation happening from time step to time step, the variation happening 
over a sub-annual time scale (if you have sub-annual data), and the pattern that 
emerges over multiple years - increases over time or long-term cycles. 

We can think about this with the following equation. 
Observed value = Trend + Seasonal Signal + Residual Variation

The goal of time series is to extract these different time scales of patterns 
so we can better understand the dynamics occurring in our data.

## Extracting a trend

The first step in time series decomposition is removing any long-term trends 
or long-term (multi-year) cycles in the data.

A moving average is a classic way of extracting the 'cross year' pattern in the 
data. What a moving average is doing is smoothing over the high frequency, or 
shorter-term fluctuations in the data so that the large-scale movements in the 
data become more apparent.

When conducting a moving average, we need to tell the computer the size of the 
window we're averaging over. This is called the order. So a MA-5 was a window 5 
time units wide. Which means for May, we are averaging values from march-july. 

One issue with moving averages is that we lose data on the front and back because 
as the name implies it is averaging over a window of values - if there is no 
march or july, it won't calculate a moving average value for any window missing 
that data.

These are typically odd so that the window is balanced.

So, let's do a moving average on this data. To do this, we're going load yet 
another package slider, which calculates the moving average values, and create 
a new column in our tsibble so that these moving averages are still associated 
with our time variable. In slider, we tell the function how many time steps 
before and after each time point we want to include in our moving average. If I 
want a ma=13 how many months before and after should I tell it?

**Answer: 6 and 6 **

```{r}
library(slider)
data_ts = data_ts |> mutate('ma_13' = slide_dbl(NDVI, mean, .before = 6, .after=6, .complete=TRUE))

```
.complete is an argument that tells slider to only calculate the moving average 
for a time point if it has data for the full window before and after the time 
point. 

These values at the beginning of the time series are missing because slider 
cannot make a complete window. 6 months before the very first time point there 
is no data. If you scroll down to the bottom of the datafile, you'll see the same 
thing. 

Now let's look at what this smoothed time series looks like, plotted over top of 
our observed monthly data
```{r}
autoplot(data_ts,NDVI) + autolayer(data_ts, ma_13, color="blue", lwd=1)
```

What the moving average does is smooths over the short-term variation in the 
time series, allowing the long-term patterns to emerge. This is the first step 
in class decomposition methods.  Larger window sizes will smooth the data further, 
but at the expense of fewer and fewer time points as the ends of the time series 
get truncated. 

The next step in time series decomposition is to remove the trend from the 
observed data to see what is left over. 


```{r}
data_ts = data_ts |> mutate("detrend" = NDVI - ma_13)
autoplot(data_ts, detrend, color="blue") + autolayer(data_ts, NDVI)
```
So we've pulled out the trend. We can see that our y-axis has been reset 

Our next step in time series decomposition is to extract the seasonal signal. 

The most basic way to extract a seasonal signal is to calculate the long-term 
average for each month. So take the January values for every year and generate 
an overall January  average and so on. Those average monthly values constitute 
our seasonal signal. To remove the seasonal signal from our data, to get the 
residual variation If we then correct out the monthly values like we did for the 
trend values, we have the signal in our data that cannot be explained by a 
long-term trend in the data or seasonal signals.

Detrended monthly data - long-term monthly averages = residual variation

Instead of doing this by hand, in the interest of time, we're going to have a 
function from the fpp3 package do it for us. We going to ask it to run a classic decomposition using our NDVI data using an additive model for how the different
time scales combine to create the observed data. There is also a multiplicative 
model: Observed = trend*season*random. You use these for timeseries that seem to 
have increasing variation with increasing mean.

the component() argument at the end extracts the output from the time series 
decomposition model for us to use more easily.

```{r}
add_decomp = data_ts |> model(classical_decomposition(NDVI, type = 'additive')) |> components()

autoplot(add_decomp)
```

We see the original time series at the top. Followed by our trend, our seasonal 
signal, and the random or residual variation. The y-axis scale changes with each 
graph. Ignore the bars they are not essential to understanding time series and 
only this package does it.

At the bottom is all the variation in the data that could not be explained by 
the other time scales. 

Notice anything odd about the seasonal signal?

**Answer:It's extremely regular**

Different approaches make different assumptions about the stability of the 
seasonal signal. Because the seasonal signal is calculated as the average value 
for each month, this approach assumes that there is no change in the pattern of 
seasonality.

## STL Decomposition ##

If we want more flexibility in our seasonal signal, we have to use a more 
complicated approach. STL decomposition (Season Trend decomposition using Loess) 
is a slightly different approach for extracting the components from the time 
series. LOESS is Locally Estimated scatterplot smoothing. Basically it is doing 
a sliding window but with regression and not averages. It allows it to be more 
flexible. The residuals from these regressions contain the seasonal and 
residual variation. To extract the seasonal signal, the STL still calculates 
the average value for each month, but instead of using all years, it uses a 
sliding window of years to calculate the January average.

**Illustrate on Board**

And again, the residual or random variation is everything that is left over.
```{r}
stl_output = data_ts |> model(STL(NDVI ~ trend(window=21) + 
                                    season(window = 13), robust = TRUE)) |> 
  components()
```
Window sizes must be odd. For trend, window size is the number of consecutive 
months of data used to fit the regression and it must be larger than your season 
window. For season this is the number of consecutive  years of a month's data to 
be used for the seasonal signal. Robust = True invokes procedures that keep STL 
from being overly influenced by extreme points when conducting the trend and 
seasonal analyses. We're using 21 and 13 here because these are actually the 
default values of the function. We'll change those in a minute. 

```{r}

autoplot(stl_output)
```
Our trend looks different in part because of the approach but also because 21 
months is much larger than the window size we were using with classic 
decomposition. The big difference though, is in our seasonal signal, where we 
now see a shift in our seasonal signal over time where our instead of 2 peaks per 
year, we seem to be moving more towards 1 very large peak in our plant 
productivity at the site. 

Let's force the stl to be more similar to our classic decomposition:

```{r}
data_ts |> model(STL(NDVI ~ trend(window=13) + season(window = "periodic"), robust = TRUE)) |> components() |> autoplot()
```
A little more movement in the trend data and our regular seasonal signal has 
returned. 

How do you choose reasonable window sizes? Frankly, it's an art not a science.
Time series decomposition is taking the inherent variation in the data and 
trying to parse out what seems to be related to long-term annual and seasonal 
patterns versus random fluctuations. As such, you generally want there to be no 
clear long-term signals in your remainder and a long-term signal sneaking 
through is often a sign that you have windows that are too large (therefore 
not picking up shorter-term long-term fluctuations). So for example, imagine a 
system where every 12 months the system flips from a productive to an 
unproductive one. In a productive year, every month is above average and during an unproductive year, every month is below average. A 21 month window is going to 
average over that to some extent, because it will combine good years and bad years. 
STL's flexible seasonal approach also won't pick that up, because it will be 
averaging 13 years of january, february, etc. So that variability will end up in 
the residuals as a signal. So, looking at your remainder graph and examining it 
for signal is the best approach to figuring out if your window sizes make sense.

For our STL, I think you could argue that the seasonal signal is leaking through 
to the residuals. Those strong peaks that are occurring with some regularity 
could be because we're not letting the seasonal signal change quickly enough to 
absorb the speed of the changing seaosnal cycle.

Let's test that:

```{r}
data_ts |> model(STL(NDVI ~ trend(window=21) +
                       season(window = 7), 
                     robust = TRUE)) |> 
  components() |> 
  autoplot()
```
And now we seem something that is a little more balanced in the remained. We 
have both troughs and peaks. The regularity and magnitude of the peaks have been 
reduced. Having some knowledge of your system and data will be useful in making 
these types of decisions. 

**Homework: In a separate R script, load the Portal data, import into a tsibble,
and decompose the rain and rodent data using the stl method on rodents and rain. 
Explain what window sizes you used for each and why**


### Video Tutorial

1. Watch Introduction to time series decomposition
<iframe width="560" height="315" src="https://www.youtube.com/embed/D45ddUVcEfY" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

2. Watch Loading data for time series decomposition in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/qVDAxVx6D_Q" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

3. Convert date column into a date format, using as.Date()

4. Watch Importing data into a time series object in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/OXrn7hI08TI" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

5. Convert the rodent column into a time series object using ts()

6. Watch Identifying the Long-term Signal in a Time Series
<iframe width="560" height="315" src="https://www.youtube.com/embed/2wcUuTqrj60" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

7. Watch Conducting a moving average in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/16ClJ5piZW4" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

8. Do a moving average with the rodent time series object using ma()

9. Watch time series decomposition: removing the long-term signal
<iframe width="560" height="315" src="https://www.youtube.com/embed/oxMZ_cVUTP0" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

10. Watch multiplicative vs. additive time series decomposition in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/iG9pOaQmvJs" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

11. Pull the trend out of the rodent time series object, using either the additive or multiplicative approach

12. Watch Using decompose() to do a time series decomposition in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/nYAQMnAguQ8" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

13. Apply decompose() to the rodent time series object


14. Watch Using Season Trend Decomposition using Loess (stl) in R
<iframe width="560" height="315" src="https://www.youtube.com/embed/Y8OHQjcYEjU" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

15. Use stl() on the rodent time series object

16. Watch Time Series Decomposition Wrap up
<iframe width="560" height="315" src="https://www.youtube.com/embed/v3dkh4YnX5w" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

17. Submit your r code (either as a file or cut and paste text) through the assignment for this module in the course canvas site

---

